Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-01-23 MScDiplom_Lemenkova.tex
corrected references to figures, added acknowledgements

\section*{Acknowledgements}
	\markright{Acknowledgements}
	\addcontentsline{toc}{section}{Acknowledgements}
\pagebreak

\section*{Disclaimer}
	\markright{Disclaimer}
	\addcontentsline{toc}{section}{Disclaimer}
		This document describes work undertaken as part of a programme of study at the University of
		Twente, Faculty ITC. All views and opinions expressed therein remain the sole responsibility of
		the author, and do not necessarily represent those of the university.\\
\pagebreak

@@

\pagenumbering{arabic}
\rfoot{MSc Thesis\\ \emph{"Seagrass mapping and monitoring along the coasts of Crete, Greece"}}
\section{Introduction}
%%%%%%%%%%%%%%%%%%Section1%%%%%%%%%%%%%%%%%%
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\pagebreak
%%%%%%%%%%%%%%%%%Section2%%%%%%%%%%%%%%%%%%%%%%
\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
\markright{II. Seagrass monitoring: overview of literature and research resources}
\renewcommand{\headrulewidth}{0.4pt}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

%%%%%%%%%%%%%%%%%  Section  3  %%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and methods}
\counterwithin{figure}{section}
\counterwithin{table}{section}

In the section \nameref{sec:2} we briefly discussed main existing cartographic methods and tools for seagrss mapping. In the current  research we try to combine various techniques from mentioned above for the assessment of the seagrass distribution in the Mediterranean environment, for the case study of Crete. We mentioned in \nameref{sec:0} the software that we used for logically separated research sections, which required various approach. Thus, for example,  spectral analysis and assessment of spectral signaturesis technically based on WASI, sttistical analysis of series of experimental raw data has been made by means of statistical software, such as \href{http://www.spss.com/}{SPSS}; mapping has been made using  \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS}; images processing and classification were made using\href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine} software by means of Google Earth and  \href{http://landsat.gsfc.nasa.gov/}{Landsat} imagery.

@@

annual mean water temperatures and geological factors, i.e. seafloor conditions and sediments. The

@@

\subsection{Study area}
General area: Island of Crete, Greece (Fig.\ref{fig:16}).\\
The study area of the current MSc research is located in the shallow areas of the Ligaria beach, Agia
Pelagia and Xerocampos, Crete island, Greece (Fig.\ref{fig:17}). These shelf areas have maximal depth of four meters.
Seagrass sampling has been performed at two stations at a depth of 4 meters, in the following selected
areas:
1. Ligaria beach (Agia Pelagia district), 36°20′N 22°59′E
2. Xerokampos, 35°12′N 26°18′E
3. Agia Pelagia, 36°20′N 22°59′E

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.10]{Fig-16.jpg}
	\caption{Study area: Crete island}
	\label{fig:18}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.25]{Fig-17.jpg}
	\caption{Study area: Ligaria beach, Crete island}
	\label{fig:19}
\end{figure}

The Ligaria Beach is a narrow, sandy and pebble beach (Fig.12), located ca 15 km north-west from Heraklion.
The in-situ measurements were conducted during the fieldwork in the period 21.09.2010-11.10.2010.
The fieldwork on seagrass monitoring included visual estimations and photo- and video footage of the
above-ground seagrass patches, sediment seafloor cover types, species compositions, water depth and
geograpic locations recorded using GPS.
\pagebreak

\subsubsection{Fieldwork equipment}
The research materials and equipment were provided by \href{http://www.nhmc.uoc.gr/index-2.htm}{the Natural History Museum of Crete}, the \href{http://www.uoc.gr/Department/index.html}{University of Crete} and the ITC, and included the following items (Fig.\ref{fig:18}): 1) three iPAQs; 2)
three GPS; 3) Three underwater video cameras, Olympus ST 8000, suitable for
photographing up to up to 33 foot depths and high-resolution 12.0-Mpixel image sensor; 4)
Markers and cords for the depths measurements; 5) Waterproof plasctic Otterbox for keeping the
iPAQ dry; 6) \href{http://www.padi.com/scuba/}{SCUBA Diving equipment}, taken from the Ligaria Diving Centre; 7) Boat (Fig.\ref{fig:19});

\begin{figure}[h]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:20a}\includegraphics[width=0.3\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:20b}\includegraphics[width=0.3\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:20c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:20d}\includegraphics[width=0.25\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment}
	\label{fig:20}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.70]{Fig-18.jpg}
	\caption{Boat used for the fieldwork measurements on Crete island. 2010}
	\label{fig:21}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements}
	\label{fig:22}
\end{figure}

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras Olympus ST 8000 (Fig.\ref{fig:18}), at
five minutes interval along the tracklog path (Fig.\ref{fig:21}).
research sampling design also included measurements of depths (Fig. \ref{fig:63}), because bathymetry is one of the
most determinating factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, iPAQs, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The GPS allowed to capture measurement locations on the iPAQ,
encapsulated in a plastic waterproof Otterbox (Fig \ref{}). Camera were adjusted horizontally by a leveller and
were mounted under the bottom of the boat to capture video footage and imagery (Fig.\ref{fig:20}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. The
locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig \ref{66}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria}
	\label{fig:23}
\end{figure}

Along the Ligaria beach there are several types of the seagrass landscapes, namely, segrass meadows
continuously covering vast areas (Fig.\ref{fig:62} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:62} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:62} (c)).
The results of the underwater videometric measurements of the Olympus cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass P.oceanica meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig. \ref{fig:65}). Seagrass species on the Ligaria and Agia
Pelagia beach consist of P.oceanica. The types of sediment on the Ligaria vary from coarse sand,
P.oceanica patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:64}).

@@

\subsection{Review of the collected data}
The remote sensing imagery were collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery include satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
	\item[] 1. Optical spectra of P.oceanica, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:23})
	\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
	\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:68})
	\item[]4. Results of underwater videometric measurements of the Olympus cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:68}) and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.

@@

The spatial resolution of Landsat ETM+ image is 30 m in the visible and near infrared bands (bands
1-5 and 7); the spatial resolution of ASTER 15 m for the visible and near-infrared bands. IKONOS
acquires data in 3 visible channels and NIR, with spatial resolution of 1-4 meters. ASTER and
IKONOS images are suggested to be included as soon as available, in addition to the Landsat. The
research includes fieldwork in-situ measurements of the seagrass distribution along the northern and
south-eastern coasts of Crete in chosen locations.\\
The optical measurements of the irradiance and radiance of the sea water and bottom cover types of
the seafloor have been received by the means of the optical sensors Trios-RAMSES Hyperspectral
UVA/UVB Irradiance Sensor and RAMSES-ARC Hyperspectral UV-VIS Radiance Sensor (Fig. \ref{fig:10}), both adjusted for the measurements of the irradiance and radiance (see Appendices: Tab. \ref{tab:18}, Tab. \ref{tab:19}).
The highest measured values are located in the diapason of 410-730 nm for the water irradiance (Fig. \ref{fig:49}) and 430-650 for the water radiance (Fig.\ref{fig:47}). 
Afterwards, the measured values of the
radiance and irradiance, respectively, have been used for the computation of the spectral reflectance
properties of the sea water and bottom cover types (Fig.\ref{fig:23}). The spectral range of radiance
cover diapason of 320-950 nm, and ittadiance measurements are covered in the interval of 280-500
nm, which is suitable for a characteristics of seagrass reflectance.
Different curves on the reflectance, radiance and irradiance graphs for example, on Fig.\ref{fig:43}, or \ref{fig:45}, enlarged) represent several series of
the measurements. The values of the spectral reflectance are received from the
computations of these values using Excel mathematical formulae. The graphs shown on Fig.\ref{fig:47}, Fig. \ref{fig:48}, Fig. \ref{fig:49} display values of the radiance and irradiance of the sea water in Agia Pelagia, with and without sediments and suspended particles, respectively. Graph on Fig \ref{fig:44} displays statistuical analysis of the measured sets of observations, i.e. shows the midspread of the statistical quartiles (Q1-Q3), mean and extreme values.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot}}
	\label{fig:25}
\end{figure}
\pagebreak

@@

A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonatic sand - consist of multiple measurement sets of P.oceanica made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October, 1-445 for 14th October for P.oceanica (Tab. \ref{tab:10}); 1-106 for water without sedimnts, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab. \ref{tab:17}). All these data have been analysed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualise distribution of the values at varuious spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while P.oceanica - at the interval 318-951 nm. We visualised the behavour of different spectra on relevant graphs (Fig. \ref{fig:43}, \ref{fig:44}, \ref{fig:45}). 

@@

potential that it may have on the approaches for further images processing and classification. Behavour of the spectra of P.oceanica have been tested using datasets for various depths: 0.5, 1.5 and 2.5 meters (Tab. \ref{tab:8}, Tab. \ref{tab:9}). The raw
initial data of measurements of the spectral reflectance have been preprocessed and statistically analysed thereafter withgraphs visualising mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:33}, \ref{fig:28}). These data

@@

Therefore, these data had to be interpolated and standardized to one

@@

detailed data by interpolating them from 3 and 4 nm step up to 2 nm (Fig.\ref{fig:24}).

@@

spectrum of constant albedo at WASI (Fig.\ref{fig:25}). 

@@

further research experiment.The results of the linear interpolation (Fig.\ref{fig:25}) demonstrate values of the

@@

column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:42}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see Appendix, Fig. \ref{fig:54}, Fig. \ref{fig:55}).
After the preliminary analysis, the measured data were visualised using GNUplot programm, which
enables fine plotting of various datasets together: Fig \ref{fig:26}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:47}, Fig. \ref{fig:49}),

@@

various interpolation techniques (Fig.\ref{fig:48}), we have choosen Bézier curves interpolation (Fig.\ref{fig:49}) which contain convex

@@

various types of sand, mud, other sediments, (Fig.\ref{fig:26})) exist in well-defined and narrow (10–20 nm)

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.30]{GNU-18.jpg}
	\caption{Multiplot showing spectral reflectance of the seawater with sediments, measured in aquarium tank, Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot} \\Two complimentray graphs below show the results of the statistical analysis.}
	\label{fig:26}
\end{figure}

>>>>

\section*{Acknowledgements}
	\markright{Acknowledgements}
	\addcontentsline{toc}{section}{Acknowledgements}

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appretiate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using Gdal technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throuout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am greatful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr.\textit{Dingtian Yang} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectances of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

\section*{Disclaimer}
	\markright{Disclaimer}
	\addcontentsline{toc}{section}{Disclaimer}
		This document describes work undertaken as part of a programme of study at the University of
		Twente, Faculty ITC. All views and opinions expressed therein remain the sole responsibility of
		the author, and do not necessarily represent those of the university.\\

\vspace{3em}

		I herewith certify that although I confered with other researchers while preparing this document, and used a range of various research and literature 			resources, cited in the text, the content of this MSc Thesis document remains my own, original work, written by me personally. \\

		Signed.................................(Lemenkova Polina) \\

		Enschede, The Netherlands. \\ 18th February 2011.
\pagebreak

@@

\pagenumbering{arabic}
\rfoot{MSc Thesis\\ \emph{"Seagrass mapping and monitoring along the coasts of Crete, Greece"}}
\section{Introduction}
%%%%%%%%%%%%%%%%%%  Section 1   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\pagebreak
%%%%%%%%%%%%%%%%%  Section2   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
\markright{II. Seagrass monitoring: overview of literature and research resources}
\renewcommand{\headrulewidth}{0.4pt}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

%%%%%%%%%%%%%%%%%  Section  3  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and methods}
\counterwithin{figure}{section}
\counterwithin{table}{section}

In the section \nameref{sec:2} we briefly discussed main existing cartographic methods and tools for seagrss mapping. In the current  research we try to combine various techniques from mentioned above for the assessment of the seagrass distribution in the Mediterranean environment, for the case study of Crete. We mentioned in \nameref{sec:0} the software that we used for logically separated research sections, which required various approach. Thus, for example,  spectral analysis and assessment of spectral signaturesis technically based on WASI, sttistical analysis of series of experimental raw data has been made by means of statistical software, such as \href{http://gretl.sourceforge.net/}{Gretl} and \href{http://www.spss.com/}{SPSS}; mapping has been made using  \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS}; images processing and classification were made using\href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine} software by means of Google Earth and  \href{http://landsat.gsfc.nasa.gov/}{Landsat} imagery.

@@

annual mean water temperatures and geological factors, i.e. seafloor conditions and sediments. \\ The

@@

\subsection{Study area}
General area: Island of Crete, Greece (Fig.\ref{fig:3.1}).\\
The study area of the current MSc research is located in the shallow areas of the Ligaria beach, Agia
Pelagia and Xerocampos, Crete island, Greece (Fig.\ref{fig:3.2}). These shelf areas have maximal depth of four meters.
Seagrass sampling has been performed at two stations at a depth of 4 meters, in the following selected
areas:
1. Ligaria beach (Agia Pelagia district), 36°20′N 22°59′E
2. Xerokampos, 35°12′N 26°18′E
3. Agia Pelagia, 36°20′N 22°59′E

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.10]{Fig-16.jpg}
	\caption{Study area: Crete island}
	\label{fig:3.1}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.25]{Fig-17.jpg}
	\caption{Study area: Ligaria beach, Crete island}
	\label{fig:3.2}
\end{figure}

The Ligaria Beach is a narrow, sandy and pebble beach (Fig.\ref{fig:3.2}), located ca 15 km north-west from Heraklion.
The in-situ measurements were conducted during the fieldwork in the period 21.09.2010-11.10.2010.
The fieldwork on seagrass monitoring included visual estimations and photo- and video footage of the
above-ground seagrass patches, sediment seafloor cover types, species compositions, water depth and
geograpic locations recorded using GPS.
\pagebreak

\subsubsection{Fieldwork equipment}
The research materials and equipment were provided by \href{http://www.nhmc.uoc.gr/index-2.htm}{the Natural History Museum of Crete}, the \href{http://www.uoc.gr/Department/index.html}{University of Crete} and the ITC, and included the following items (Fig.\ref{fig:3.3}): 1) three iPAQs (Fig.\ref{fig:3.3a}); 2)
three GPS (Fig.\ref{fig:3.3c}); 3) Three underwater video cameras, Olympus ST 8000  (Fig.\ref{fig:3.3d}), suitable for
photographing up to up to 33 foot depths and high-resolution 12.0-Mpixel image sensor; 4)
Markers and cords for the depths measurements; 5) Waterproof plasctic Otterbox  (Fig.\ref{fig:3.3b})for keeping the
iPAQ dry; 6) \href{http://www.padi.com/scuba/}{SCUBA Diving equipment}, taken from the Ligaria Diving Centre; 7) Boat (Fig.\ref{fig:3.4});

\begin{figure}[h]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:3.3a}\includegraphics[width=0.3\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:3.3b}\includegraphics[width=0.3\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:3.3c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:3.3d}\includegraphics[width=0.25\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment}
	\label{fig:3.3}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.70]{Fig-18.jpg}
	\caption{Boat used for the fieldwork measurements on Crete island. 2010}
	\label{fig:3.4}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements}
	\label{fig:3.5}
\end{figure}

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras Olympus ST 8000  (Fig.\ref{fig:3.3d}), at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determinating factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, iPAQs, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The GPS allowed to capture measurement locations on the iPAQ,
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). Camera were adjusted horizontally by a leveller and
were mounted under the bottom of the boat to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. The
locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig \ref{fig:A.7}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria}
	\label{fig:3.6}
\end{figure}

Along the Ligaria beach there are several types of the seagrass landscapes, namely, segrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes,
which located separately from broad meadows (Fig.\ref{fig:A.3} (c)).
The results of the underwater videometric measurements of the Olympus cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass P.oceanica meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig. \ref{fig:A5}). Seagrass species on the Ligaria and Agia
Pelagia beach consist of P.oceanica. The types of sediment on the Ligaria vary from coarse sand,
P.oceanica patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

@@

\subsection{Review of the collected data}
The remote sensing imagery were collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery include satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
	\item[] 1. Optical spectra of P.oceanica, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:A.23})
	\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
	\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:A.37})
	\item[]4. Results of underwater videometric measurements of the Olympus cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:A.38}) and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.

@@

The spatial resolution of Landsat ETM+ image is 30 m in the visible and near infrared bands (bands
1-5 and 7); the spatial resolution of ASTER 15 m for the visible and near-infrared bands. IKONOS
acquires data in 3 visible channels and NIR, with spatial resolution of 1-4 meters. ASTER and
IKONOS images are suggested to be included as soon as available, in addition to the Landsat. The
research includes fieldwork in-situ measurements of the seagrass distribution along the northern and
south-eastern coasts of Crete in chosen locations.\\
The optical measurements of the irradiance and radiance of the sea water and bottom cover types of
the seafloor have been received by the means of the optical sensors Trios-RAMSES Hyperspectral
UVA/UVB Irradiance Sensor and RAMSES-ARC Hyperspectral UV-VIS Radiance Sensor (Fig. \ref{fig:2.1}), both adjusted for the measurements of the irradiance and radiance (see Appendices: Tab. \ref{tab:A.1}, Tab. \ref{tab:A.2}).
The highest measured values are located in the diapason of 410-730 nm for the water irradiance (Fig. \ref{fig:A.18}) and 430-650 for the water radiance (Fig.\ref{fig:A.16}). 
Afterwards, the measured values of the
radiance and irradiance, respectively, have been used for the computation of the spectral reflectance
properties of the sea water and bottom cover types (Fig.\ref{fig:A.15}). The spectral range of radiance
cover diapason of 320-950 nm, and ittadiance measurements are covered in the interval of 280-500
nm, which is suitable for a characteristics of seagrass reflectance.
Different curves on the reflectance, radiance and irradiance graphs for example, on Fig.\ref{fig:A.12}, or \ref{fig:A.14}, enlarged) represent several series of
the measurements. The values of the spectral reflectance are received from the
computations of these values using mathematical formulae. The graphs shown on Fig. \ref{fig:4.2}, Fig. \ref{fig:A.23}, Fig. \ref{fig:A.24}, Fig. \ref{fig:A.17} display values of the radiance and irradiance of the sea water in Agia Pelagia, with and without sediments and suspended particles, respectively. Graph on Fig \ref{fig:A.13} displays statistuical analysis of the measured sets of observations, i.e. shows the midspread of the statistical quartiles (Q1-Q3), mean and extreme values.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{GNU-17.jpg}
	\caption{Optical properties of the sea water with sediments, measured in aquarium tank. Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}				{Gnuplot}}\label{fig:4.2}
\end{figure}
\pagebreak

@@

A collection of visible spectra of two seafloor cover types - \textit{P.oceanica} and carbonatic sand - consist of multiple measurement sets of P.oceanica made by means of RAMSES hyperspectral radiometer: 1-350 for 15th October (fragment on Tab. \ref{tab:A.8}), 1-445 for 14th October for P.oceanica; 1-106 for water without sedimnts, measured in aquarium tank, 1-27 for seawater with sediments measured in aquarium tank; 1-75 for carbonate sand (Tab. \ref{tab:17}). All these data have been analysed and statistically tested. We measured mean, extreme values (min-max), median and statistical quartiles, in order to visualise distribution of the values at varuious spectral wavelengths. Carbonate sand was measured at wavelengths 402-750 nm, while P.oceanica - at the interval 318-951 nm. We visualised the behavour of different spectra on relevant graphs (Fig. \ref{fig:4.4}, \ref{fig:A.35}, \ref{fig:A.28}). 

@@

potential that it may have on the approaches for further images processing and classification. Behavour of the spectra of P.oceanica have been tested using datasets for various depths: 0.5, 1.5 and 2.5 meters (Tab. \ref{tab:8}, Tab. \ref{tab:A.7}). The raw
initial data of measurements of the spectral reflectance have been preprocessed and statistically analysed thereafter with graphs visualising mean spectra and Q1-Q3 interval, instead of the series of single observations which is illustrated, for example, on Figures \ref{fig:27}, \ref{fig:4.10}, \ref{fig:28}). These data

@@

Therefore, these data had to be interpolated (Fig. \ref{fig:A.10}) and standardized to one

@@

detailed data by interpolating them from 3 and 4 nm step up to 2 nm (Fig.\ref{fig:A.8}).

@@

spectrum of constant albedo at WASI (Fig.\ref{fig:A.9}). 

@@

further research experiment.The results of the linear interpolation (Fig.\ref{fig:A.35}) demonstrate values of the

@@

column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see Appendix, Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualised using Gnuplot programm, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),

@@

various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have choosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contain convex

@@

various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})) exist in well-defined and narrow (10–20 nm)

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.30]{GNU-18.jpg}
	\caption{Multiplot showing spectral reflectance of the seawater with sediments, measured in aquarium tank, Agia Pelagia district, Crete. \href{http://www.gnuplot.info/}{Gnuplot} \\Two complimentray graphs below show the results of the statistical analysis.}
	\label{fig:4.4}
\end{figure}
